---
title: "Converting between YAML and R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Converting between YAML and R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(zuyaml)
```

`zuyaml` converts between YAML 1.2 and ordinary R objects. This vignette
documents the conversion rules, which are part of the package's contract: they
are meant to be predictable rather than clever, and every place where
information is lost is listed at the end.

## The API

Eight functions, on one rule: **`_all` means "YAML stream"**.

|            | one document     | stream of documents  |
|------------|------------------|----------------------|
| parse text | `yaml_parse()`   | `yaml_parse_all()`   |
| emit text  | `yaml_emit()`    | `yaml_emit_all()`    |
| read file  | `yaml_read()`    | `yaml_read_all()`    |
| write file | `yaml_write()`   | `yaml_write_all()`   |

A YAML *stream* holds several documents separated by `---`. It is a different
thing from a *sequence*, and the API keeps them apart: `yaml_parse()` refuses
anything but exactly one document, so trailing documents are never silently
discarded.

```{r}
yaml_parse("a: 1")

yaml_parse_all("---\na: 1\n---\nb: 2\n")

try(yaml_parse("---\na: 1\n---\nb: 2\n"))
```

An empty stream is an error rather than `NULL`, because `NULL` is the
legitimate result of parsing a document whose content is `null`.

## Scalars

Scalars follow the YAML 1.2 **core schema**. The most visible consequence is
that `yes` and `no` are strings, not booleans — that was a YAML 1.1 rule.

```{r}
str(yaml_parse("[null, true, 42, 3.14, hello, yes, 2026-09-07]"))
```

Dates stay strings: turning them into `Date` would be policy beyond a parser's
job. Quoted scalars are always strings, so quoting is how you protect a value
that would otherwise look like a number.

```{r}
str(yaml_parse('[42, "42"]'))
```

`.inf` and `.nan` map to `Inf` and `NaN`, and an empty *quoted* string stays a
string rather than becoming `NULL`.

```{r}
str(yaml_parse('["", ~, .inf]'))
```

### Large integers

R has no exact integer type beyond 2^53. Rather than round silently, values
past that point become a `zuyaml_bigint`, a character vector holding the
decimal value.

```{r}
x <- yaml_parse("9223372036854775807")
x
class(x)
```

It carries the decimal normalisation, not the source text, so hexadecimal input
compares equal to its decimal spelling:

```{r}
as.character(yaml_parse("0x7FFFFFFFFFFFFFFF"))
```

Use `big_integers` to choose a different policy — `"double"` opts in to losing
precision, `"error"` refuses the document.

```{r}
yaml_parse("9223372036854775807", big_integers = "double")
```

## Sequences and mappings

A sequence becomes a list, and a mapping with scalar keys becomes a named list.

```{r}
str(yaml_parse("- 1\n- two\n- true\n"))
str(yaml_parse("host: localhost\nport: 8080\n"))
```

`simplify = TRUE` collapses sequences whose elements are all scalars of one
type. It is **not** the default, because it makes the shape of the result
depend on the contents of the document: `[1, 2]` would give an integer vector
while `[1, "a"]` gives a list, so code that indexes the result would be correct
only for the inputs its author happened to test.

```{r}
str(yaml_parse("[1, 2, 3]"))
str(yaml_parse("[1, 2, 3]", simplify = TRUE))
```

### Duplicate keys

Rejected by default, because the resulting object would be ambiguous to index.

```{r}
try(yaml_parse("a: 1\na: 2\n"))
```

Non-string scalar keys are stringified, so `1` and `"1"` would collide as R
names. That collision is caught too rather than quietly producing duplicate
names.

```{r}
str(yaml_parse("1: a\n2: b\n"))
try(yaml_parse('1: a\n"1": b\n'))
```

### Keys that are not scalars

YAML allows a sequence or mapping to be a key. Such a mapping cannot become a
named list without destroying structure, so it becomes a `zuyaml_map`: two
parallel lists, `keys` and `values`.

```{r}
yaml_parse("? [one, two]\n: value\n")
```

## Anchors and aliases

Aliases are resolved to the value of their target. Node identity is not
preserved — R has no portable way to express it.

```{r}
str(yaml_parse("defaults: &d\n  timeout: 30\nserver: *d\n"))
```

Cycles are refused, and `aliases = "error"` rejects aliases outright.

## Emitting

```{r}
cat(yaml_emit(list(host = "localhost", port = 8080L, tls = TRUE)))
```

Strings that would resolve as another type under the core schema are quoted, so
they survive a round trip:

```{r}
cat(yaml_emit(list(version = "42", ratio = 0.5, name = "true")))
```

Ambiguous input is refused rather than guessed at. A partially named list has
no obvious meaning as either a sequence or a mapping:

```{r}
try(yaml_emit(list(a = 1, 2)))
try(yaml_emit(data.frame(a = 1)))
```

`list()` cannot express both an empty sequence and an empty mapping, so it
emits as `[]`, and an empty *named* list is how you ask for `{}`.

```{r}
cat(yaml_emit(list()))
cat(yaml_emit(structure(list(), names = character())))
```

## Untrusted input

Three limits apply, all enforced by `zuyaml` itself:

- `max_size` — input bytes;
- `max_depth` — nesting depth;
- `max_nodes` — total values materialised.

The third exists because the first two do not bound *alias expansion*. A
billion-laughs document is small and shallow by construction, yet expands
exponentially:

```{r}
bomb <- paste0(
  "a: &a [x,x,x,x,x,x,x,x,x]\n",
  "b: &b [*a,*a,*a,*a,*a,*a,*a,*a,*a]\n",
  "c: &c [*b,*b,*b,*b,*b,*b,*b,*b,*b]\n",
  "d: &d [*c,*c,*c,*c,*c,*c,*c,*c,*c]\n",
  "e: [*d,*d,*d,*d,*d,*d,*d,*d,*d]\n"
)
nchar(bomb)
try(yaml_parse(bomb, max_nodes = 1000))
```

## Errors

Failures are classed conditions carrying a stable code and a position, so
callers never have to match on message text.

```{r}
err <- tryCatch(yaml_parse("a: [1, 2"), zuyaml_error = identity)
err$code
err$line
err$column
```

## What does not round-trip

Every entry below is a deliberate trade-off, not an oversight.

| Construct | Behaviour | Round-trips? |
|---|---|---|
| Comments | Discarded | No |
| Quoting and block style | Re-chosen on emit | Semantically |
| Anchors and aliases | Resolved to values | Semantically |
| Non-string scalar keys | Stringified (`1` → `"1"`) | Semantically |
| Collection-valued keys | `zuyaml_map` on parse | **No** — cannot be emitted |
| Duplicate keys (opt-in) | Duplicate R names | **No** — cannot be emitted |
| `NA` | Emitted as `null`, parses back as `NULL` | **No** |
| `list(1L)` vs `c(1L)` | Both emit `- 1` | **No** |
| Whole-numbered doubles | Emit as `42`, parse back as integer | Numerically |
| `big_integers = "double"` | Precision lost | No (opt-in) |

`zuyaml` is a semantic parser, not a round-trip editor: it converts values, and
formatting is not part of the value. Preserving comments and layout would need
a different API built on a retained document.
